---
title: "Feline Serum Metabolome Stability PLOS ONE"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align="center")
library(knitr)
library(tidyverse)
library(plyr)
library(ggplot2)
library(kableExtra)
library(pheatmap)
library(fgsea)
library(factoextra)
library(EnhancedVolcano)
library(omu)
library(rstatix)
library(tidyr)
library(corrr)
library(RColorBrewer)
library(viridis)
library(visdat)
library(purrr)
library(ggpubr)
library(qvalue)
library(FactoMineR)
library(reshape2)
library(ggplus)
library(Matrix)
library(lme4)
library(lmerTest)
library(emmeans)
library(randomForest)
require(caTools)
library(stringr)
library(performance)

```

# **LOAD DATA**

```{r}
metab <- read.csv("Feline_serumMetab_Stability_Metabolites.csv", row.names = 1)
  
annot <- read.csv("Feline_serumMetab_Stability_Annot.csv", row.names = 1)
  
meta <- read.csv("Feline_serumMetab_Stability_MetaData.csv", row.names = 1)
```

```{r}
#log-transform for statistical analysis

metab_norm <- log(metab)

```

# **EXPLORE DATA**

How many serum metabolites were detected?

```{r}
nrow(annot)
```

Metabolites were measured across 8 super pathways. Examining the distribution of different metabolic super-pathways among the named metabolites, we find that lipids and amino acids are the most prevalent.

```{r}
table(annot$SUPER.PATHWAY)

#length(unique(annot$SUPER.PATHWAY))
```


```{r}

SuperPath <- as.data.frame(table(annot$SUPER.PATHWAY))

names(SuperPath)[1] <- "SuperPathway"

SuperPath <- SuperPath[order(-SuperPath$Freq), ]

path.col <- viridis_pal()(nrow(SuperPath))

names(path.col) <- SuperPath$SuperPathway

super_plot <- ggplot(data = SuperPath, aes(y = Freq, x = reorder(SuperPathway, Freq), 
                                           fill = SuperPathway)) + 
  geom_bar(stat = "identity") + 
  scale_fill_manual("Legend", values = path.col) +
  xlab("") +
  ylab("Number of Detected Metabolites") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position = "none")  +
  theme(axis.text.x=element_text(angle=45,hjust=1)) +
  coord_flip() 

super_plot + ggtitle("Number of Metabolites in Each Super-Pathway")


```

86 sub-pathways were identified, here are the top 50:

```{r}

SubPath <- as.data.frame(table(annot$SUB.PATHWAY))

names(SubPath)[1] <- "SubPathway"

SubPath <- SubPath[order(-SubPath$Freq), ]

SubPath_top20 <- SubPath[1:50, ]

SubPath_top20$SuperPath <- annot$SUPER.PATHWAY[match(SubPath_top20$SubPathway, annot$SUB.PATHWAY)]

sub_plot <- ggplot(data = SubPath_top20, aes(y = Freq, x = reorder(SubPathway, Freq), 
                                               fill = SuperPath)) + 
  scale_fill_manual("Legend", values = path.col) +
  geom_bar(stat = "identity") + 
  labs(fill = "") +
  xlab("") +
  ylab("Number of Detected Metabolites") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.text.x=element_text(angle=45,hjust=1)) +
  theme(text = element_text(size = 8))+
  coord_flip() 

sub_plot + ggtitle("Metabolites in the Top 50 Sub-Pathways")

```
# **UNSUPERVISED ANALYSIS**

## **PRINCIPAL COMPONENT ANALYSIS**

```{r, include=F}

#make new df
dat_pca <- data.frame(t(metab_norm))

meta$condition <- paste(meta$Temp, meta$Time, sep = "_")

dat_pca$condition <- meta$condition[match(rownames(dat_pca), rownames(meta))]

dat_pca$ID <- meta$ID[match(rownames(dat_pca), rownames(meta))]

dat_pca$condition <- as.factor(dat_pca$condition)
dat_pca$ID <- as.factor(dat_pca$ID)


dat_pca <- dat_pca[, c(735, 734, 1:733)]

#perform PCA
pca <- PCA(dat_pca[, 3:734], graph = F, scale.unit = T)


```


```{r}

fviz_eig(pca)

```

```{r}
dat_pca$condition <- factor(dat_pca$condition, 
                              levels = c("-80C_12 Months", "-20C_6 Months", "-20C_12 Months"))

#make a beautiful plot using ggplot2

dat_pca$pc1 <- pca$ind$coord[, 1] # indexing the first column

dat_pca$pc2 <- pca$ind$coord[, 2]  # indexing the second column

pca.vars <- pca$var$coord %>% data.frame

pca.vars$vars <- rownames(pca.vars)

pca.vars.m <- melt(pca.vars, id.vars = "vars")

pca_plot <- ggplot(data = dat_pca, aes(x = pc1, y = pc2, color = ID)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  guides(color = "none") +
  geom_point(aes(shape = condition, color = ID), size = 3) +
  xlab("PC 1 (23.94%)") + 
  ylab("PC 2 (17.67%)") +
  theme_minimal() +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  theme(legend.position="bottom") + 
  theme(legend.title=element_text(size=12), 
    legend.text=element_text(size=12)) +
  labs(color = "Cat ID", shape = "Storage Condition")

pca_plot

```

## **HEATMAP OF EUCLIDIAN DISTANCES**

```{r}

meta$new_ID <- gsub("_", "", meta$ID)

meta$new_ID <- paste(meta$new_ID, gsub(" Months", "", meta$condition), sep = "_")

annotation_colors <- list(Condition =
                            c(`-80C_12 Months` = "red",
                              `-20C_6 Months` = "blue",
                              `-20C_12 Months` = "purple"))

#annotation_colors <- list(Condition = c("red", "blue", "purple"))

annotation_col <- as.data.frame(meta[, 5])

names(annotation_col)[1] <- "Condition"

rownames(annotation_col) <- meta$new_ID

#I get en error about infinite values when making the heatmap, this gets rid of analytes with zero variance
metab_heat <- metab_norm[apply(metab_norm, MARGIN = 1, FUN = function(x) sd(x) != 0),]

table(rownames(meta) == names(metab_heat))

names(metab_heat) <- meta$new_ID

heat_plot <- pheatmap(
  mat               = metab_heat,
  color             = inferno(10),
  border_color      = NA,
  show_colnames     = T,
  show_rownames     = FALSE,
  annotation_col    = annotation_col,
  annotation_colors = annotation_colors,
  cluster_cols = T,
  clustering_distance_cols = "euclidean",
  drop_levels       = TRUE,
  scale             = "row",
  fontsize          = 6,
  fontsize_col = 6,
  annotation_legend = T,
  treeheight_row = 0
)

heat_plot

```
# **LINEAR MIXED EFFECTS MODELS OF RELATIVE METABOLITE CONCENTRATIONS**

```{r, include=F}

metab_lm <- data.frame(t(metab_norm))

#add sample data

metab_lm$condition <- meta$condition[match(rownames(metab_lm), rownames(meta))]

metab_lm$ID <- meta$ID[match(rownames(metab_lm), rownames(meta))]

metab_lm <- metab_lm[, c(735, 734, 1:733)]

metab_lm$condition <- factor(metab_lm$condition,
                              levels = c("-80C_12 Months", "-20C_6 Months", "-20C_12 Months"))

metab_lm$ID <- factor(metab_lm$ID)

```

Loop over dataframe to perform linear mixed effects models:

```{r, inlcude = F}
# names of variables
names <- colnames(metab_lm)[3:734]
nNames <- length(names)

lm.res <- NULL

lm_pvals.res <- NULL

for (i in names) {
  # print status
  print(paste("Running entity:", i, "which is", which(names==i), "out of", nNames))
  name <- i
  f <- formula(paste(i, "~ condition + (1|ID)"))
  fit_test <- lmer(f, data=metab_lm)
  
  fit_df <- data.frame(summary(fit_test)[10])
  fit_df$feature <- name
  
  fit_df_pval <- data.frame(anova(fit_test)[5:6])
  fit_df_pval$feature <- name

  lm.res <- data.frame(rbind(lm.res, fit_df))
  lm_pvals.res <- data.frame(rbind(lm_pvals.res, fit_df_pval))
  
}

```

```{r}
#calculate FDR-adjusted p-values to control false discovery
lm_pvals.res$FDR <- qvalue(lm_pvals.res$Pr..F.)$qvalues

lm_pvals.res$Biochemical <- annot$BIOCHEMICAL[match(lm_pvals.res$feature, rownames(annot))]

lm_pvals.res$SuperPathway <- annot$SUPER.PATHWAY[match(lm_pvals.res$feature, rownames(annot))]

lm_pvals.res$SubPathway <- annot$SUB.PATHWAY[match(lm_pvals.res$feature, rownames(annot))]

rownames(lm_pvals.res) <- lm_pvals.res$feature

lm_pvals.res <- lm_pvals.res[, c(5:7, 1:2, 4)]

names(lm_pvals.res)[5] <- "P.value"

sig_lm.res <- lm_pvals.res[lm_pvals.res$FDR < 0.05, ]

#output results
write.csv(lm_pvals.res, "lmer_res.csv")

```

How many metabolite features varied significantly by storage condition?

```{r}
nrow(sig_lm.res)

```

Assess pairwise contrasts for significantly variable metabolites:

```{r, inlcude}
# names of variables
names <- rownames(sig_lm.res)
nNames <- length(names)

lm_contrasts.res <- NULL

for (i in names) {
  # print status
  print(paste("Running entity:", i, "which is", which(names==i), "out of", nNames))
  name <- i
  f <- formula(paste(i, "~ condition + (1|ID)"))
  fit <- lmer(f, data = metab_lm)
  
  contrast_df <- data.frame(contrast(emmeans(fit, ~condition), interaction = "pairwise",
                                     adjust = "fdr"))
  contrast_df$feature <- name
  
  lm_contrasts.res <- data.frame(rbind(lm_contrasts.res, contrast_df))
  
}

lm_contrasts.res$Biochemical <- annot$BIOCHEMICAL[match(lm_contrasts.res$feature, rownames(annot))]

lm_contrasts.res$SuperPathway <- annot$SUPER.PATHWAY[match(lm_contrasts.res$feature, rownames(annot))]

lm_contrasts.res$SubPathway <- annot$SUB.PATHWAY[match(lm_contrasts.res$feature, rownames(annot))]

lm_contrasts.res <- lm_contrasts.res[, c(8:10, 1:6)]

sig_lm_contrasts.res <- lm_contrasts.res[lm_contrasts.res$p.value < 0.05, ]

#output results
write.csv(lm_contrasts.res, file = "lm_contrasts.csv")

```

How many pairwise contrasts were significant?

```{r}

nrow(sig_lm_contrasts.res)

```

## **PLOT SIGNIFICANTLY VARIABLE METABOLITES**

Prepare data and set up contrasts:

```{r, include=F}

sig_names <- rownames(sig_lm.res)

sig_metab <- data.frame(t(metab_norm), check.names = F)

# #for some reason there are NAs in this vector, remove
# sig_names<-sig_names[!sig_names == "NA.1"]
# sig_names<-sig_names[!sig_names == "NA"]

sig_metab <- sig_metab %>% select(sig_names)

table(names(sig_metab) == sig_names)

names(sig_metab) <- annot$BIOCHEMICAL[match(names(sig_metab), rownames(annot))] 

sig_metab$Condition <- meta$condition[match(rownames(sig_metab), rownames(meta))] 

sig_metab$Condition <- factor(sig_metab$Condition, 
                              levels = c("-80C_12 Months", "-20C_6 Months", "-20C_12 Months"))


sig_metab$Condition <- revalue(sig_metab$Condition, 
                               c("-80C_12 Months"="-80C_12", 
                                 "-20C_6 Months"="-20C_6",
                                 "-20C_12 Months"="-20C_12"))



cols <- c("red", "blue", "purple")

my_comparisons <- list(c("-80C_12", "-20C_6"), 
                        c("-80C_12", "-20C_12"), 
                        c("-20C_6", "-20C_12"))

```


```{r}

base_plot <-sig_metab %>%
  gather(Measure, Value, -Condition) %>%
  ggplot(aes(x = factor(Condition), y = Value, fill = Condition)) +
  geom_boxplot() +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme(legend.title = element_text(size = 8), 
        legend.text = element_text(size = 8),
        axis.text=element_text(size=8),
        axis.title=element_text(size=8)) +
  scale_fill_manual(values = cols) +
  facet_wrap(~Measure, 
             scales = "free") +
  xlab("") +
  ylab("Relative Metabolite Concentration (Log, Median-Scaled)") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position="bottom") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) + 
  theme(strip.text.x = element_text(
        size = 8))


pdf("S_Fig_sigBoxplots.pdf")
facet_multiple(base_plot, "Measure",
               nrow = ,
               ncol = 2,
               scales = "free")
dev.off()


```

# **INTRACLASS CORRELATION ANALYSIS**

 - ICC < 0.5: Poor stability across storage conditions  
 - ICC 0.5 - 0.74: Fair stability across storage conditions  
 - ICC 0.7 - 0.9: Moderate stability across storage conditions  
 - ICC > 0.9: Excellent stability across storage conditions 

```{r, include=F, warning=F, message=F}

names <- colnames(metab_lm)[3:734]
nNames <- length(names)

icc.res <- NULL

for (i in names) {
  # print status
  #print(paste("Running entity:", i, "which is", which(names==i), "out of", nNames))
  name <- i
  f <- formula(paste(i, "~ condition + (1|ID)"))
  fit <- lmer(f, data = metab_lm)
  
  icc <-as.data.frame(try(performance::icc(fit)))[1]
  names(icc) <- "Conditional_ICC"
  feature <- i
  names(feature) <- "feature"
  icc <- data.frame(cbind(feature, icc))

  icc.res <- rbind(icc, icc.res)
  
}

icc.res$Biochemical <- annot$BIOCHEMICAL[match(icc.res$feature, rownames(annot))] 
 
icc.res$SubPathway <- annot$SUB.PATHWAY[match(icc.res$feature, rownames(annot))] 

icc.res$SuperPathway <- annot$SUPER.PATHWAY[match(icc.res$feature, rownames(annot))] 

```

```{r, include=F}
icc.res$Stability <- case_when(
  icc.res$Conditional_ICC >= 0.9 ~ "Excellent (ICC≥0.9)",
  icc.res$Conditional_ICC >= 0.75 & icc.res$Conditional_ICC < 0.9 ~ "Moderate (ICC 0.75-0.89)",
  icc.res$Conditional_ICC >=0.5 & icc.res$Conditional_ICC < 0.75 ~ "Fair (ICC 0.5-0.74)",
  icc.res$Conditional_ICC < 0.5 ~ "Poor (ICC≤0.5)"
)


stability <- data.frame(table(icc.res$Stability))
stability
names(stability)[1] <- "Metabolite Stability"

stability$Prop <- stability$Freq/nrow(icc.res)

stability$Prop <- round(stability$Prop, 2)

rownames(icc.res) <- icc.res$feature

icc.res <- icc.res[, c(3:5, 2, 6)]

names(icc.res)[4] <- "ICC"

#output results
write.csv(icc.res, "ICC_res.csv")

```

```{r}
stability
```

```{r}
stability$`Metabolite Stability` <- factor(stability$`Metabolite Stability`, 
                                           levels = c("Poor (ICC≤0.5)", 
                                                      "Fair (ICC 0.5-0.74)", 
                                                      "Moderate (ICC 0.75-0.89)", 
                                                      "Excellent (ICC≥0.9)"))



stability_plot <- ggplot(data = stability, aes(y = Freq, x = `Metabolite Stability`)) + 
  geom_bar(stat = "identity", color = "black", fill = "lightblue") + 
  labs(fill = "") +
  xlab("Metabolite Stability") +
  ylab("Number of Metabolites") + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position = "none")  +
  theme(axis.text.x=element_text(angle=45,hjust=1)) +
  geom_text(aes(label = paste(Prop, "%", sep = "")), size = 3, nudge_y = 12) +
  coord_flip()

stability_plot

```

# **METABOLITE SET ENRICHMENT ANALYSIS**

Create pre-ranked vector of metabolites based on the ICC:

```{r}

diff_enrich <- icc.res[, c(1, 2, 4)]

diff_enrich <- diff_enrich[order(diff_enrich$ICC, decreasing = T), ]

rank <- diff_enrich[, 3]
names(rank) <- diff_enrich$Biochemical

pathways <- diff_enrich[, c(1, 2)]

pathways <- pathways[order(pathways$SubPathway, pathways$Biochemical),]

pathways$SubPathway <- as.character(pathways$SubPathway)

pathways <- split(pathways$Biochemical, as.character(pathways$SubPathway))

rank <- rank[!is.na(rank) ==T]

```

Run MSEA:


```{r, include=F, warning=F, message=F}

set.seed(321) #for reproducability
fgseaRes <- fgsea(pathways = pathways, 
                  stats = rank,
                  minSize=2,
                  maxSize=500
                  )

fgseaRes$SuperPathway <- annot$SUPER.PATHWAY[match(fgseaRes$pathway, annot$SUB.PATHWAY)]

fgseaRes_out <- data.frame(fgseaRes)

fgseaRes_out <- apply(fgseaRes_out,2,as.character)

fgseaRes_out <- fgseaRes_out[, c(1,9,2:8)]

write.csv(fgseaRes_out, file = "fgseaRes.csv")

```

NOTE: this is a permutational test, results vary across runs

PLot results:

```{r}

topPathwaysUp <- fgseaRes[ES > 0 & pval < 0.05][head(order(pval), n=20), pathway]

topPathwaysDown <- fgseaRes[ES < 0 & pval < 0.05][head(order(pval), n=10), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))


topPathways <- unlist(topPathways)

fgseaRes_top20 <- fgseaRes[fgseaRes$pathway %in% topPathways, ]

fgseaRes_top20 <- fgseaRes_top20[order(fgseaRes_top20$NES), ]

fgseaRes_top20$pathway <- as.factor(fgseaRes_top20$pathway)

fgseaRes_top20$pathway <- factor(fgseaRes_top20$pathway, 
                                 levels = fgseaRes_top20$pathway[order(fgseaRes_top20$NES)])

gsea_plot <- ggplot(data = fgseaRes_top20, aes(y = fgseaRes_top20$NES, 
                                               x = fgseaRes_top20$pathway, 
                                               fill = ifelse(fgseaRes_top20$padj 
                                                             <0.05, 
                                                             "Padj<0.05", "Padj≥0.05"))) +
  geom_bar(stat = "identity", position = "dodge") + ylim(-4, 4) + coord_flip()

gsea_plot <- gsea_plot + 
  labs(fill = "") +
  xlab("") +
  ylab("Normalized Enrichment Score") + scale_fill_manual(values=c("orange", "grey")) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position = "right") + theme(legend.title.align=0.5)
      
gsea_plot <- gsea_plot + theme(axis.text = element_text(size = 6), 
                               legend.position = "bottom")

gsea_plot


```


