---
title: "Feline Metabolome Stability Study"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align="center")
library(knitr)
library(tidyverse)
library(plyr)
library(ggplot2)
library(kableExtra)
library(pheatmap)
library(fgsea)
library(factoextra)
library(EnhancedVolcano)
library(omu)
library(rstatix)
library(tidyr)
library(corrr)
library(RColorBrewer)
library(viridis)
library(visdat)
library(purrr)
library(ggpubr)
#library(made4)
library(qvalue)
library(FactoMineR)
library(reshape2)
library(ggplus)
library(Matrix)
library(lme4)
library(lmerTest)
library(emmeans)
#library(ICC)
library(randomForest)
require(caTools)
```

#**Data Summary**

These data were generated from an experiment designed to assess the stability of serum metabolites exposed to sub-optimal storage conditions.
Samples of serum from 8 healthy cats from a research colony were aliqoted and stored at -20C for 6 months, -20C for 12 months, and -80C for 12 months. After the storage period, samples kept at -20C were transferred to -80C. 

Untargeted quantification of serum metabolites was performed by Metabolon Inc.

```{r, inlcude=F}

#load scaled, imputed metabolite abundance data

metab <- read.csv("Feline_serumMetab_Stability_Metabolites.csv", header = T, row.names = 1)

#load sample metadata

meta <- read.csv("Feline_serumMetab_Stability_MetaData.csv", header = T, row.names = 1)

#load metabolite annotation data

annot <- read.csv("Feline_serumMetab_Stability_Annot.csv", header = T, row.names = 1)

#str(meta)

meta$Temp <- factor(meta$Temp, levels = c("-80C", "-20C"))

meta$Time <- factor(meta$Time)

rownames(meta) <- names(metab)

#sanity check

table(rownames(meta) == names(metab))

sum(rownames(annot) == rownames(metab))

```

```{r}
#list.files()

table(annot$PLATFORM)
```


```{r}

#log-transform

metab_log <- log(metab)

```


How many serum metabolites were detected?

```{r}
nrow(annot)
```

Metabolites were measured across 8 super pathways. Examining the distribution of different metabolic super-pathways among the named metabolites, we find that lipids and amino acids are the most prevalent.

```{r}
table(annot$SUPER.PATHWAY)

#length(unique(annot$SUPER.PATHWAY))
```


```{r}

SuperPath <- as.data.frame(table(annot$SUPER.PATHWAY))

names(SuperPath)[1] <- "SuperPathway"

SuperPath <- SuperPath[order(-SuperPath$Freq), ]

path.col <- viridis_pal()(nrow(SuperPath))

names(path.col) <- SuperPath$SuperPathway

super_plot <- ggplot(data = SuperPath, aes(y = Freq, x = reorder(SuperPathway, Freq), 
                                           fill = SuperPathway)) + 
  geom_bar(stat = "identity") + 
  scale_fill_manual("Legend", values = path.col) +
  xlab("") +
  ylab("Number of Detected Metabolites") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position = "none")  +
  theme(axis.text.x=element_text(angle=45,hjust=1)) +
  coord_flip() 

super_plot + ggtitle("Number of Metabolites in Each Super-Pathway")


```

86 sub-pathways were identified, here are the top 50:

```{r}

#length(unique(annot$SUB.PATHWAY))

SubPath <- as.data.frame(table(annot$SUB.PATHWAY))

length(unique(SubPath$SubPathway))

names(SubPath)[1] <- "SubPathway"

SubPath <- SubPath[order(-SubPath$Freq), ]

SubPath_top20 <- SubPath[1:50, ]

SubPath_top20$SuperPath <- annot$SUPER.PATHWAY[match(SubPath_top20$SubPathway, annot$SUB.PATHWAY)]

sub_plot <- ggplot(data = SubPath_top20, aes(y = Freq, x = reorder(SubPathway, Freq), 
                                               fill = SuperPath)) + 
  scale_fill_manual("Legend", values = path.col) +
  geom_bar(stat = "identity") + 
  labs(fill = "") +
  xlab("") +
  ylab("Number of Detected Metabolites") +
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(axis.text.x=element_text(angle=45,hjust=1)) +
  theme(text = element_text(size = 8))+
  coord_flip() 

sub_plot + ggtitle("Metabolites in the Top 50 Sub-Pathways")

```

# **Unsupervised Analysis**

## **Principal Component Analysis**

```{r, include=F}

#make new df
dat_pca <- data.frame(t(metab_log))

meta$condition <- paste(meta$Temp, meta$Time, sep = "_")

dat_pca$condition <- meta$condition[match(rownames(dat_pca), rownames(meta))]

dat_pca$ID <- meta$ID[match(rownames(dat_pca), rownames(meta))]

dat_pca$condition <- as.factor(dat_pca$condition)
dat_pca$ID <- as.factor(dat_pca$ID)


dat_pca <- dat_pca[, c(735, 734, 1:733)]

#perform PCA
pca <- PCA(dat_pca[, 3:734], graph = F, scale.unit = T)

#summary(pca)

fviz_eig(pca)

summary(pca)
```

```{r}

dat_pca$condition <- case_when(dat_pca$condition == "-80C_12 Months"~ "-80°C 12 Months",
                               dat_pca$condition == "-20C_6 Months"~ "-20°C 6 Months",
                               dat_pca$condition == "-20C_12 Months"~ "-20°C 12 Months")
  
dat_pca$condition <- factor(dat_pca$condition, 
                              levels = c("-80°C 12 Months", "-20°C 6 Months", "-20°C 12 Months"))

dat_pca$pc1 <- pca$ind$coord[, 1] # indexing the first column

dat_pca$pc2 <- pca$ind$coord[, 2]  # indexing the second column

pca.vars <- pca$var$coord %>% data.frame

pca.vars$vars <- rownames(pca.vars)

pca.vars.m <- melt(pca.vars, id.vars = "vars")

pca_plot <- ggplot(data = dat_pca, aes(x = pc1, y = pc2, color = ID)) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2) +
  guides(color = "none") +
  geom_point(aes(shape = condition, color = ID), size = 3) +
  xlab("PC 1 (23.94%)") + 
  ylab("PC 2 (17.67%)") +
  theme_minimal() +
  theme(panel.grid = element_blank(), 
        panel.border = element_rect(fill= "transparent")) +
  theme(legend.position="bottom") + 
  theme(legend.title=element_text(size=12), 
    legend.text=element_text(size=12)) +
  labs(color = "Cat ID", shape = "Storage Condition")

pca_plot

```


## **Heatmap of Metabolite Abundance**

```{r}

meta$new_ID <- gsub("_", "", meta$ID)

meta$new_ID <- paste(meta$new_ID, gsub(" Months", "", meta$condition), sep = "_")

#extract hex color codes for default ggplot2 colors to sync annotation colors for catID
hex <- hue_pal()(8)

annotation_colors <- list(Condition =
                            c(`-80C_12 Months` = "red",
                              `-20C_6 Months` = "blue",
                              `-20C_12 Months` = "purple"),
                          CatID = 
                            c(Cat_1 = hex[1],
                              Cat_2 = hex[2],
                              Cat_3 = hex[3],
                              Cat_4 = hex[4],
                              Cat_5 = hex[5],
                              Cat_6 = hex[6],
                              Cat_7 = hex[7],
                              Cat_8 = hex[8])
                          )

#annotation_colors <- list(Condition = c("red", "blue", "purple"))

annotation_col <- as.data.frame(meta[, 5])

names(annotation_col)[1] <- "Condition"

rownames(annotation_col) <- meta$new_ID

ID <- c(rep(1:8), rep(1:8), rep(1:8))

ID <- paste("Cat", ID, sep = "_")

annotation_col$CatID <- ID

annotation_col$Condition

annotation_col$Condition <- case_when(annotation_col$Condition == "-80C_12 Months"~ "-80°C 12 Months",
                               annotation_col$Condition == "-20C_6 Months"~ "-20°C 6 Months",
                               annotation_col$Condition == "-20C_12 Months"~ "-20°C 12 Months")

names(annotation_colors$Condition) <- case_when(names(annotation_colors$Condition) == "-80C_12 Months"~ "-80°C 12 Months",
                               names(annotation_colors$Condition) == "-20C_6 Months"~ "-20°C 6 Months",
                               names(annotation_colors$Condition) == "-20C_12 Months"~ "-20°C 12 Months")
  
# dat_pca$condition <- factor(dat_pca$condition, 
#                               levels = c("-80°C 12 Months", "-20°C 6 Months", "-20°C 12 Months"))

#I get en error about infinite values when making the heatmap, this gets rid of it
metab_heat <- metab_log[apply(metab_log, MARGIN = 1, FUN = function(x) sd(x) != 0),]

table(rownames(meta) == names(metab_heat))

names(metab_heat) <- meta$new_ID

heat_plot <- pheatmap(
  mat               = metab_heat,
  color             = inferno(10),
  border_color      = NA,
  show_colnames     = F,
  show_rownames     = FALSE,
  annotation_col    = annotation_col,
  annotation_colors = annotation_colors,
  cluster_cols = T,
  clustering_distance_cols = "euclidean",
  drop_levels       = TRUE,
  scale             = "row",
  fontsize          = 6,
  fontsize_col = 6,
  annotation_legend = T,
  treeheight_row = 0
)

heat_plot

```

# **Hypothesis Testing**

Linear mixed effect models were used to detect significant differences in serum metabolite abundance among the three storage conditions. The metabolites abundance data were previously median-centered and log-transformed.


```{r, include=F}

metab_lm <- data.frame(t(metab_log))

#add sample data

metab_lm$condition <- meta$condition[match(rownames(metab_lm), rownames(meta))]

metab_lm$ID <- meta$ID[match(rownames(metab_lm), rownames(meta))]

metab_lm <- metab_lm[, c(735, 734, 1:733)]

metab_lm$condition <- factor(metab_lm$condition,
                              levels = c("-80C_12 Months", "-20C_6 Months", "-20C_12 Months"))

metab_lm$ID <- factor(metab_lm$ID)

#met421 causes problems in the linear model - remove it

metab_lm <- metab_lm[, !names(metab_lm )== "met421"]
```



```{r, inlcude = F}
# names of variables
names <- colnames(metab_lm)[3:734]
nNames <- length(names)

lm.res <- NULL

lm_pvals.res <- NULL

for (i in names) {
  # print status
  print(paste("Running entity:", i, "which is", which(names==i), "out of", nNames))
  name <- i
  f <- formula(paste(i, "~ condition + (1|ID)"))
  fit_test <- lmer(f, data=metab_lm)
  
  fit_df <- data.frame(summary(fit_test)[10])
  fit_df$feature <- name
  
  fit_df_pval <- data.frame(anova(fit_test)[5:6])
  fit_df_pval$feature <- name

  lm.res <- data.frame(rbind(lm.res, fit_df))
  lm_pvals.res <- data.frame(rbind(lm_pvals.res, fit_df_pval))
  
}


```

```{r}
lm_pvals.res$FDR <- qvalue(lm_pvals.res$Pr..F.)$qvalues

lm_pvals.res$Biochemical <- annot$BIOCHEMICAL[match(lm_pvals.res$feature, rownames(annot))]

lm_pvals.res$SuperPathway <- annot$SUPER.PATHWAY[match(lm_pvals.res$feature, rownames(annot))]

lm_pvals.res$SubPathway <- annot$SUB.PATHWAY[match(lm_pvals.res$feature, rownames(annot))]

rownames(lm_pvals.res) <- lm_pvals.res$feature

lm_pvals.res <- lm_pvals.res[, c(5:7, 1:2, 4)]

names(lm_pvals.res)[5] <- "P.value"

sig_lm.res <- lm_pvals.res[lm_pvals.res$FDR < 0.05, ]

nrow(sig_lm.res)

names_sig <- rownames(sig_lm.res)

write.csv(lm_pvals.res, "lmer_res.csv")

```

Pairwise contrasts will be performed for all metabolites. 

```{r, inlcude}
# names of variables
names <- colnames(metab_lm)[3:734]
nNames <- length(names)

lm_contrasts_all.res <- NULL

for (i in names) {
  # print status
  print(paste("Running entity:", i, "which is", which(names==i), "out of", nNames))
  name <- i
  f <- formula(paste(i, "~ condition + (1|ID)"))
  fit <- lmer(f, data = metab_lm)
  
  contrast_df <- data.frame(contrast(emmeans(fit, ~condition), interaction = "pairwise",
                                     adjust = "fdr"))
  contrast_df$feature <- name
  
  lm_contrasts_all.res <- data.frame(rbind(lm_contrasts_all.res, contrast_df))
  
}

lm_contrasts_all.res$Biochemical <- annot$BIOCHEMICAL[match(lm_contrasts_all.res$feature, rownames(annot))]

lm_contrasts_all.res$SuperPathway <- annot$SUPER.PATHWAY[match(lm_contrasts_all.res$feature, rownames(annot))]

lm_contrasts_all.res$SubPathway <- annot$SUB.PATHWAY[match(lm_contrasts_all.res$feature, rownames(annot))]

```

Calculate log2 Fold-Changes

```{r, include=F}

metab_means <- metab %>% t() %>% data.frame() %>%
  mutate_if(is.numeric, log2) %>%
  mutate(Group=meta$condition) %>% 
  group_by(Group) %>%
   dplyr::summarise(across(
    .cols = is.numeric, 
    .fns = list(Mean = mean), 
    na.rm = T, 
    .names = "{col}_{fn}"
    ))

metab_means <- metab_means[c(3,2,1), ]

names <- colnames(metab_means)[2:734]
nNames <- length(names)

foldchange.res <- NULL

for (i in names) {
  # print status
  #print(paste("Running entity:", i, "which is", which(names==i), "out of", nNames))
  name <- i
   
  fc1 <- data.frame(metab_means[2,i] - metab_means[1,i])
  fc1$contrast <- "minus20C_6months_vs_minus80C"
  names(fc1)[1] <- "Log2FC"
  fc1$met <- i
  
  fc2 <- data.frame(metab_means[3,i] - metab_means[1,i])
  fc2$contrast <- "minus20C_12months_vs_minus80C"
  names(fc2)[1] <- "Log2FC"
  fc2$met <- i
  
  fc3 <- data.frame(metab_means[3,i] - metab_means[2,i])
  fc3$contrast <- "minus20C_12months_vs_minus20C_6months"
  names(fc3)[1] <- "Log2FC"
  fc3$met <- i
  
  foldchange.res <- rbind(foldchange.res, fc1, fc2, fc3)

}

foldchange.res$met <- gsub("_Mean", "", foldchange.res$met)

```


```{r}

keep <- unique(lm_contrasts_all.res$feature)

foldchange.res <- foldchange.res[foldchange.res$met %in% keep, ]

all_contrast.res <- cbind(foldchange.res, lm_contrasts_all.res)

all_contrast.res <- all_contrast.res[, c(10:11, 13, 12, 2, 5:9, 1)]

names(all_contrast.res)[10] <- c("q.value")
```

**Volcano Plots**

How many metabolites differed between -20C/12 months and -80C?

```{r}
volcano1 <- all_contrast.res[all_contrast.res$contrast %in% "minus20C_12months_vs_minus80C", ]

sig_res1 <- volcano1[volcano1$q.value < 0.05, ]

nrow(sig_res1)
```

```{r}
V1 <- ggplot(volcano1, aes(x=Log2FC, y=-log10(q.value))) + 
  geom_point() + 
  theme_bw() +
  geom_hline(yintercept=-log10(0.05), col="red") +
  labs(y= "-log(q-value)", x = "Fold Change (log2)") +
  ggtitle("-20°C 12 Months vs. -80°C") +
  theme(plot.title = element_text(size = 10))

V1
```


How many metabolites differed between -20C/6 months and -80C?


```{r}
volcano2 <- all_contrast.res[all_contrast.res$contrast %in% "minus20C_6months_vs_minus80C", ]

sig_res2 <- volcano2[volcano2$q.value < 0.05, ]

nrow(sig_res2)

```

```{r}
V2 <-ggplot(volcano2, aes(x=Log2FC, y=-log10(q.value))) + 
  geom_point() + 
  theme_bw() +
  geom_hline(yintercept=-log10(0.05), col="red") +
  labs(y= "-log(q-value)", x = "Fold Change (log2)") +
  ggtitle("-20°C 6 Months vs. -80°C") +
  theme(plot.title = element_text(size = 10))

V2
```

How many metabolites differed between -20C/12 months and -20C/6 months?

```{r}
volcano3 <- all_contrast.res[all_contrast.res$contrast %in% "minus20C_12months_vs_minus20C_6months", ]

sig_res3 <- volcano3[volcano3$q.value < 0.05, ]

nrow(sig_res3)

```

```{r}
V3 <-ggplot(volcano3, aes(x=Log2FC, y=-log10(q.value))) + 
  geom_point() + 
  theme_bw() +
  geom_hline(yintercept=-log10(0.05), col="red") +
  labs(y= "-log(q-value)", x = "Fold Change (log2)") +
  ggtitle("-20°C 12 Months vs. -20°C 6 Months") +
  theme(plot.title = element_text(size = 10))

V3
```

```{r}
all_sig_res <- rbind(sig_res1, sig_res2, sig_res3)

```


For the purposes of generating results with the lowest possible chances of false discovery, only pairwise contrasts from metabolites that varied significantly in the overall linear models will be considered statistically significant.

```{r}
names_sig

all_sig_res <- all_sig_res[all_sig_res$feature %in% names_sig, ]

length(unique(all_sig_res$feature))

write.csv(all_sig_res, "Fel_MetabStability_posthoc.csv")

```


```{r, include=F}

sig_metab <- data.frame(t(metab_log), check.names = F)

# #for some reason there are NAs in this vector, remove
# sig_names<-sig_names[!sig_names == "NA.1"]
# sig_names<-sig_names[!sig_names == "NA"]

sig_metab <- sig_metab[, names(sig_metab) %in% names_sig]

table(names(sig_metab) == names_sig)

names(sig_metab) <- annot$BIOCHEMICAL[match(names(sig_metab), rownames(annot))] 

sig_metab$Condition <- meta$condition[match(rownames(sig_metab), rownames(meta))] 

sig_metab$Condition <- case_when(sig_metab$Condition == "-80C_12 Months"~ "-80°C 12 Months",
                               sig_metab$Condition == "-20C_6 Months"~ "-20°C 6 Months",
                               sig_metab$Condition == "-20C_12 Months"~ "-20°C 12 Months")

sig_metab$Condition <- factor(sig_metab$Condition,
                              levels = c("-80°C 12 Months", "-20°C 6 Months", "-20°C 12 Months"))
# 
# 
# sig_metab$Condition <- revalue(sig_metab$Condition, 
#                                c("-80C_12 Months"="-80C_12", 
#                                  "-20C_6 Months"="-20C_6",
#                                  "-20C_12 Months"="-20C_12"))



cols <- c("red", "blue", "purple")

my_comparisons <- list(c("-80°C 12 Months", "-20°C 6 Months"), 
                        c("-80°C 12 Months", "-20°C 12 Months"), 
                        c("-20°C 6 Months", "-20°C 12 Months"))

```


```{r}

library(stringr)

keep <- unique(sig_lm_contrasts.res$feature)

sig_lm_contrasts2 <- lm_contrasts.res[lm_contrasts.res$feature 
                                          %in% keep,]

string <- str_split(sig_lm_contrasts2$condition_pairwise, "\\) - \\(")

x <- data.frame(string)

x <- data.frame(t(x))

sig_lm_contrasts2$group1 <- x$X1

sig_lm_contrasts2$group2 <- x$X2

sig_lm_contrasts2 <- sig_lm_contrasts2[, c(1:2, 6)]

# ggplot(sig_metab, aes(x = Condition, y = sig_metab$`12,13-DiHOME`)) + 
#   geom_boxplot()

```

```{r}

base_plot <-sig_metab %>%
  gather(Measure, Value, -Condition) %>%
  ggplot(aes(x = factor(Condition), y = Value, fill = Condition)) +
  geom_boxplot() +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  theme(legend.title = element_text(size = 8), 
        legend.text = element_text(size = 8),
        axis.text=element_text(size=8),
        axis.title=element_text(size=8)) +
  scale_fill_manual(values = cols) +
  facet_wrap(~Measure, 
             scales = "free") +
  xlab("") +
  ylab("Relative Metabolite Concentration (Log, Median-Scaled)") +
  stat_compare_means(comparisons = my_comparisons, 
                     label = "p.signif", 
                     method = "t.test", 
                     p.adjust.method = "BH", 
                     hide.ns = T,
                     vjust = 0.5) +  
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position="bottom") +
  theme(axis.text=element_text(size=6)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) + 
  theme(strip.text.x = element_text(
        size = 8))


pdf("S_Fig_sigBoxplots.pdf")
facet_multiple(base_plot, "Measure",
               nrow = ,
               ncol = 2,
               scales = "free")
dev.off()


```



```{r}
# glutathione and cysteine metabolites

sig_metab_glut <- sig_metab

sig_metab_glut <- sig_metab_glut[, c(25:27, 46, 76)]

glut_plot <-sig_metab_glut %>%
  gather(Measure, Value, -Condition) %>%
  ggplot(aes(x = factor(Condition), y = Value, fill = Condition)) +
  geom_boxplot() +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  scale_fill_manual(values = cols) +
  facet_wrap(~Measure, 
             scales = "free") +
  xlab("") +
  ylab("Relative Concentration (Log, Median-Scaled)") +
  stat_compare_means(comparisons = my_comparisons, 
                     label = "p.signif", 
                     method = "t.test", 
                     p.adjust.method = "BH", 
                     hide.ns = T,
                     vjust = 0.5) +  
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position="bottom") +
  theme(legend.title = element_text(size = 8), 
        legend.text = element_text(size = 8),
        axis.text=element_text(size=6),
        axis.title=element_text(size=8)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) 


glut_plot

```

```{r}
# amino acids

sig_metab_aa <- sig_metab

sig_metab_aa <- sig_metab_aa[, c(19, 44, 35, 37, 39, 43, 12, 76)]

amino_plot <-sig_metab_aa %>%
  gather(Measure, Value, -Condition) %>%
  ggplot(aes(x = factor(Condition), y = Value, fill = Condition)) +
  geom_boxplot() +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  scale_fill_manual(values = cols) +
  facet_wrap(~Measure, 
             scales = "free") +
  xlab("") +
  ylab("Relative Concentration (Log, Median-Scaled)") +
  stat_compare_means(comparisons = my_comparisons, 
                     label = "p.signif", 
                     method = "t.test", 
                     p.adjust.method = "BH", 
                     hide.ns = T,
                     vjust = 0.5) +  
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position="bottom") +
  theme(legend.title = element_text(size = 8), 
        legend.text = element_text(size = 8),
        axis.text=element_text(size=5),
        axis.title=element_text(size=8)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) + 
  theme(strip.text.x = element_text(
        size = 8))

amino_plot

```



```{r}

sig_metab_PUFA <- sig_metab

sig_metab_PUFA <- sig_metab_PUFA[, c(16, 30:33, 51, 76)]

PUFA_plot <-sig_metab_PUFA %>%
  gather(Measure, Value, -Condition) %>%
  ggplot(aes(x = factor(Condition), y = Value, fill = Condition)) +
  geom_boxplot() +
  guides(color = guide_legend(override.aes = list(size = 2))) +
  scale_fill_manual(values = cols) +
  facet_wrap(~Measure, 
             scales = "free") +
  xlab("") +
  ylab("Relative Concentration (Log, Median-Scaled)") +
  stat_compare_means(comparisons = my_comparisons, 
                     label = "p.signif", 
                     method = "t.test", 
                     p.adjust.method = "BH", 
                     hide.ns = T,
                     vjust = 0.5) +  
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position="bottom") +
  theme(legend.title = element_text(size = 8), 
        legend.text = element_text(size = 8),
        axis.text=element_text(size=5),
        axis.title=element_text(size=8)) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) + 
  theme(strip.text.x = element_text(
        size = 8))

PUFA_plot

```



# **Intraclass Correlation Analysis**

Intraclass correlation (ICC) analysis is used to quantify the degree to which samples with a fixed degree of relatedness (e.g. same storage condition) resemble each other in terms of a quantitative trait. I will use to to detect metabolites that vary significantly among samples that were collected from the same cats but exposed to different srtorage conditions. 

Metabolites with a ICC approaching 1 have very litte variance among different strage conditions. I will use the following scheme to rank metabolites based on ICC:

 - ICC < 0.5: Poor stability across storage conditions  
 - ICC 0.5 - 0.74: Fair stability across storage conditions  
 - ICC 0.7 - 0.9: Moderate stability across storage conditions  
 - ICC > 0.9: Excellent stability across storage conditions  
 - A negative ICC will be interpreted as measures of the same metabolite over time being less similar than any two metabolites chosen randomly of the entire data.  
 
 Other investigators have used a similar approach:

 - https://doi.org/10.1371/journal.pone.0021103
 - https://doi.org/10.1371/journal.pone.0218549
 - https://doi.org/10.1038/s41598-020-72914-7
 - https://cebp.aacrjournals.org/content/cebp/25/11/1483.full.pdf


```{r}

names <- colnames(metab_lm)[3:734]
nNames <- length(names)

icc.res <- NULL

for (i in names) {
  # print status
  #print(paste("Running entity:", i, "which is", which(names==i), "out of", nNames))
  name <- i
  f <- formula(paste(i, "~ condition + (1|ID)"))
  fit <- lmer(f, data = metab_lm)
  
  icc <-as.data.frame(try(performance::icc(fit)))[1]
  names(icc) <- "Conditional_ICC"
  feature <- i
  names(feature) <- "feature"
  icc <- data.frame(cbind(feature, icc))

  icc.res <- rbind(icc, icc.res)
  
}

icc.res$Biochemical <- annot$BIOCHEMICAL[match(icc.res$feature, rownames(annot))] 
 
icc.res$SubPathway <- annot$SUB.PATHWAY[match(icc.res$feature, rownames(annot))] 

icc.res$SuperPathway <- annot$SUPER.PATHWAY[match(icc.res$feature, rownames(annot))] 

```



```{r}
icc.res$Stability <- case_when(
  icc.res$Conditional_ICC >= 0.9 ~ "Excellent (ICC≥0.9)",
  icc.res$Conditional_ICC >= 0.75 & icc.res$Conditional_ICC < 0.9 ~ "Moderate (ICC 0.75-0.89)",
  icc.res$Conditional_ICC >=0.5 & icc.res$Conditional_ICC < 0.75 ~ "Fair (ICC 0.5-0.74)",
  icc.res$Conditional_ICC < 0.5 ~ "Poor (ICC≤0.5)"
)

#how many ICC's were NA?
table(is.na(icc.res$ICC))

#get rid of NAs
#icc.res <- icc.res[!is.na(icc.res$ICC), ]

stability <- data.frame(table(icc.res$Stability))
stability
names(stability)[1] <- "Metabolite Stability"

nrow(icc.res)


stability$Prop <- stability$Freq/nrow(icc.res)

stability$Prop <- round(stability$Prop, 2)


stability

rownames(icc.res) <- icc.res$feature

icc.res <- icc.res[, c(3:5, 2, 6)]

names(icc.res)[4] <- "ICC"

write.csv(icc.res, "ICC_res.csv")

```

```{r}
hist(icc.res$ICC)

shapiro_test(icc.res$ICC)

median(icc.res$ICC, na.rm = T)

min(icc.res$ICC, na.rm = T)

max(icc.res$ICC, na.rm = T)

q1q3(icc.res$ICC)
```


```{r}
icc.res <- icc.res[with(icc.res,order(-ICC)),]


```

10 most stable metabolites:
```{r}
icc.res$Biochemical[1:10]

```

10 least stable metabolites
```{r}

icc.res$Biochemical[723:732]
```


```{r}
stability$`Metabolite Stability` <- factor(stability$`Metabolite Stability`, 
                                           levels = c("Poor (ICC≤0.5)", 
                                                      "Fair (ICC 0.5-0.74)", 
                                                      "Moderate (ICC 0.75-0.89)", 
                                                      "Excellent (ICC≥0.9)"))



stability_plot <- ggplot(data = stability, aes(y = Freq, x = `Metabolite Stability`)) + 
  geom_bar(stat = "identity", color = "black", fill = "lightblue") + 
  labs(fill = "") +
  xlab("Metabolite Stability") +
  ylab("Number of Metabolites") + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position = "none")  +
  theme(axis.text.x=element_text(angle=45,hjust=1)) +
  geom_text(aes(label = paste(Prop, "%", sep = "")), size = 3, nudge_y = 12) +
  coord_flip()

stability_plot

```

# **Enrichment Analysis**

## **Differentially Abundant Metabolites**

Using the ICC results to identify metabolite subpathways enriched in highly stable and highly unstable metabolites. 

```{r}
# icc.res$Biochemical <- annot$BIOCHEMICAL[match(icc.res$feature, rownames(annot))] 
# 
# icc.res$SubPathway <- annot$SUB.PATHWAY[match(icc.res$feature, rownames(annot))] 
# 
# icc.res$SuperPathway <- annot$SUPER.PATHWAY[match(icc.res$feature, rownames(annot))] 

diff_enrich <- icc.res[, c(1, 2, 4)]

diff_enrich <- diff_enrich[order(diff_enrich$ICC, decreasing = T), ]

rank <- diff_enrich[, 3]
names(rank) <- diff_enrich$Biochemical

pathways <- diff_enrich[, c(1, 2)]

pathways <- pathways[order(pathways$SubPathway, pathways$Biochemical),]

#pathways$Biochemical <- as.character(pathways$Biochemical)
pathways$SubPathway <- as.character(pathways$SubPathway)


pathways <- split(pathways$Biochemical, as.character(pathways$SubPathway))

rank <- rank[!is.na(rank) ==T]

```


```{r}

set.seed(123) #for reproducability
fgseaRes <- fgsea(pathways = pathways, 
                  stats = rank,
                  minSize=2,
                  maxSize=500
                  )

fgseaRes$SuperPathway <- annot$SUPER.PATHWAY[match(fgseaRes$pathway, annot$SUB.PATHWAY)]

fgseaRes_out <- data.frame(fgseaRes)

fgseaRes_out <- apply(fgseaRes_out,2,as.character)

fgseaRes_out <- fgseaRes_out[, c(1,9,2:8)]

write.csv(fgseaRes_out, file = "fgseaRes.csv")

```


```{r}

topPathwaysUp <- fgseaRes[ES > 0 & pval < 0.05][head(order(pval), n=20), pathway]

topPathwaysDown <- fgseaRes[ES < 0 & pval < 0.05][head(order(pval), n=10), pathway]
topPathways <- c(topPathwaysUp, rev(topPathwaysDown))


topPathways <- unlist(topPathways)

fgseaRes_top20 <- fgseaRes[fgseaRes$pathway %in% topPathways, ]

fgseaRes_top20 <- fgseaRes_top20[order(fgseaRes_top20$NES), ]

fgseaRes_top20$pathway <- as.factor(fgseaRes_top20$pathway)

fgseaRes_top20$pathway <- factor(fgseaRes_top20$pathway, 
                                 levels = fgseaRes_top20$pathway[order(fgseaRes_top20$NES)])

gsea_plot <- ggplot(data = fgseaRes_top20, aes(y = fgseaRes_top20$NES, 
                                               x = fgseaRes_top20$pathway, 
                                               fill = ifelse(fgseaRes_top20$padj 
                                                             <0.05, 
                                                             "q-value < 0.05", "q-value ≥ 0.05"))) +
  geom_bar(stat = "identity", position = "dodge") + ylim(-4, 4) + coord_flip()

gsea_plot <- gsea_plot + 
  labs(fill = "") +
  xlab("") +
  ylab("Normalized Enrichment Score") + 
    guides(fill = guide_legend(override.aes = list(size = 0.5)))+
  scale_fill_manual(values=c("orange", "grey")) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  theme(legend.position = "bottom") + 
  theme(legend.title.align=0.5)
      
gsea_plot <- gsea_plot +
  theme(legend.title = element_text(size = 8), 
        legend.text = element_text(size = 6),
        axis.text=element_text(size=6),
        axis.title=element_text(size=8))

gsea_plot


```

